Rotation of Reynolds stress tensor [for shear-alignment]

### 2021-03-03 write/manuscript.tex
reworked w-profile to eliminate sharp transition in surface vicinity for high Reynolds number

\documentclass[a4paper,11pt]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{natbib}

\newcommand{\p}{\partial}
\newcommand{\gzm}{{\xi}}%{\gamma z_\text{match}}}

@@

\begin{abstract}
  An analytical formulation for the profiles of stream- and span-wise velocity in turbulent Ekman flow
  is developed based on asymptotic theory as well as data from direct numerical simulation and large-eddy simulation.
  %
  The profile of the stream-wise component follows the classical viscous, logarithmic and wake scaling.
  %
  The span-wise component poses a conceptual challenge to the channel-flow analogy
  in the context of asymptotic mixing, and it exhibits a mixed scaling in the surface layer, but follows
  outer scaling for most of the outer layer.
  %
  In both profiles, the wake region can be described by smooth error-function transition from a logarithmic layer
  to the constant outer velocity.

\end{abstract}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Introduction}

The exact veering of the wind away from its geostrophic direction is due to direct and indirect (due to turbulence) frictional effects,
and it plays a role for various practical problems: 
%
It sets the amount of cross-isobaric volume flow and is thus a key factor in the life-cycle of large-scale synoptic systems. This mechanism is commonly
referred to as 'Ekman pumping' referring to the perturbation of the geostrophic equilibrium that causes the veering as it was first described
and solved for analytically by \cite{Ekman:1905tl}.
%
On the mesoscale, shear in the surface layer is known to be crucial for the initiation and sustainment of strong convective events (REFERENCE?).
% 
And locally, directional sheer of the wind in the upper part of the surface layer may cause a yaw for large wind power generation devices (REFERENCE?).
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Formulation}
%
\paragraph{Physical Problem.} We consider incompressible, turbulent Ekman flow, i.e. the turbulent flow over a flat rotating plate, as
a physical model for the truly neutral atmospheric boundary layer (ABL).
%
That is, the f-plane approximation is used confining effects of rotation to the horizontal components of
velocity and neglecting dynamical effects due to latitudinal changes of the rate of rotation. 
%
%%PAR
%
\paragraph{Notation.} The dimensional velocity vector of the numerical simulations is $\underline{U} = (U_1,U_2,U_3) = (U,V,W)$ over the coordinate system
$Oxyz$, where an approximate alignment with the surface shear stress is achieved.
%
For analysis of the results, we use two coordinate systems that are
(i)  exactly aligned with the surface shear stress and labelled by an upper index $\alpha$ as in $\underline{U}^\alpha$ for the velocity vector, and
(ii) the coordinate system aligned with the free-atmosphere geostrophic wind labelled by an upper index $G$ as in $\underline{U}^G$.
%
We denote by $u_\star$ the modulus surface friction velocity and let $Z_\star=1/u_\star$ for brevity.

\section{Towards a Universal Velocity Profile for the Turbulent Ekman Layer}
\subsection{Drag-Law} 
The drag law for Ekman flow, including prediction of surface veering, has received considerable attention in the past.
\begin{itemize}
\item \cite{Rossby:1935wa} first approach;
\item \cite{Tennekes:1973jh, Blackadar:1968ew} asymptotic similarity;
\item \cite{Spalart:1989p2436, Spalart:2008p2432, Spalart:2009p2466} seminal LES studies in search for a universal log-law;
\item \cite{Ansorge:2014hf, Ansorge:2019di} re-evaluation of the constants with geophysical focus;
\item \cite{Hogstrom:1988wg, Hgstrm:1996kt} (re-)consideration of observational data.
  %

\end{itemize}
\begin{figure}
  \centerline{\includegraphics[width=0.7\textwidth]{../plot/ustar_alpha.pdf}}
  \caption{Variation of $Z_\star$ and surface veering with Reynolds number according to
    the theory by Spalart et al. (1989) and as estimated from DNS data}
  \label{fig:drag_law}
\end{figure} 
%
Attempts were also undertaken to obtain profiles of the wind speed: 
\cite{Gryning:2007dy} present an extension of the wind-speed profile beyond the surface layer using a neutral referece profile
and a stability correction;
\cite{Kelly:2010ev}, based on a probabilistic representation of stratification, develop a model for the long-term mean windspeed
in the atmospheric boundary layer and compare this with observation at different sites; 
\cite{Kelly:2016bs} demonstrate the effect of such improved model for wind-energy applications.
%
These investigations merely focus on the wind speed and thus do not regard the actual vector, i.e. the veering of
the wind with height is not described and there is little knowledge on the profile of the span-wise velocity component
and the precise shape of the hodograph in the limit of a truly neutral Ekman boundary layer.
%
While this is indeed an academic problem, the Ekman boundary layer often serves as limit for stratified boundary layers
and thus not only needs to be understood but also well-described as reference for higher-order approaches that take into
account possible effects of stratification, roughness or other physical complications of the geophysical system.
%
\subsection{Stream-wise Velocity} 
For the stream-wise velocity profile (that in non-rotating flows due to the geometry is always aligned with the surface shear stress),
well-established theories exist for various regimes according to their distance from the wall and the relative role of
viscosity, turbulence and interaction with the outer region of the flow with the logarithmic law for the mean velocity as a central anchor point:
\begin{subequations}
  \begin{itemize}
  \item In immediate vicinity to the surface, local turbulent mixing cannot occur for the no-slip/no-penetration boundary condition,
    and the mean velocity is described by a viscous profile of the form
    \begin{equation}
      (U,W)^{\alpha+}  = (z^+,0)  
    \end{equation}
    where the direction of 
    in absence of roughness elements are marginal linear regime / viscous sub-layer \cite{Foken:2002gv,Foken:1978bc} 
  \item buffer layer transition from linear regime to log-law  
  \item log-layer (K\'arm\'an, Prandtl, Jimenez, Marusic)
  \item outer log-layer () 
  \item wake region
  \item outer region 
  \end{itemize}
\end{subequations} 
%
\begin{figure}
  \includegraphics[width=\textwidth]{../plot/u_profile.pdf}
  \caption{Shear-aligned profiles of velocity components $U^{\alpha+}$ in inner (left) and outer (right) units.} 
\end{figure}

>>>>

\documentclass[a4paper,11pt]{article}

\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{natbib}
\usepackage{xcolor}

\newcommand{\p}{\partial}
\newcommand{\gzm}{{\xi}}%{\gamma z_\text{match}}}
\newcommand{\todo}[1]{\textcolor{red}{$[$#1$]$}}

@@

\begin{abstract}
  An analytical formulation for the profiles of stream- and span-wise velocity in turbulent Ekman flow
  is developed based on asymptotic theory as well as data from direct numerical simulation and large-eddy simulation.
  %
  The profile of the stream-wise component follows the classical viscous, logarithmic and wake scaling.
  %
  The span-wise component poses a conceptual challenge to the channel-flow analogy
  in the context of asymptotic mixing, and it exhibits a mixed scaling in the surface layer, but follows
  outer scaling for most of the outer layer.
  %
  In the outer layer, the velocity component profiles can be described by an Ekman-spiral with adapted
  boundary conditions that result in a reduction of the spiral-like rotation. 
  %
\end{abstract}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Introduction}

The exact veering of the wind away from its geostrophic direction is due to direct and indirect (due to turbulence) frictional effects,
and it plays a role for various practical problems: 
%
It sets the amount of cross-isobaric volume flow and is thus a key factor in the life-cycle of large-scale synoptic systems. This mechanism is commonly
referred to as 'Ekman pumping' referring to the perturbation of the geostrophic equilibrium that causes the veering as it was first described
and solved for analytically by \cite{Ekman:1905tl}.
%
On the mesoscale, shear in the surface layer is known to be crucial for the initiation and sustainment of strong convective events (REFERENCE?).
% 
And locally, directional sheer of the wind in the upper part of the surface layer may cause a yaw for large wind power generation devices (REFERENCE?).

The drag law for Ekman flow, including prediction of surface veering, has received considerable attention in the past.
\begin{itemize}
\item \cite{Rossby:1935wa} first approach;
\item \cite{Tennekes:1973jh, Blackadar:1968ew} asymptotic similarity;
\item \cite{Spalart:1989p2436, Spalart:2008p2432, Spalart:2009p2466} seminal LES studies in search for a universal log-law;
\item \cite{Ansorge:2014hf, Ansorge:2019di} re-evaluation of the constants with geophysical focus;
\item \cite{Hogstrom:1988wg, Hgstrm:1996kt} (re-)consideration of observational data.
  %
\end{itemize} 
%
Attempts were also undertaken to obtain profiles of the wind speed: 
\cite{Gryning:2007dy} present an extension of the wind-speed profile beyond the surface layer using a neutral referece profile
and a stability correction;
\cite{Kelly:2010ev}, based on a probabilistic representation of stratification, develop a model for the long-term mean windspeed
in the atmospheric boundary layer and compare this with observation at different sites; 
\cite{Kelly:2016bs} demonstrate the effect of such improved model for wind-energy applications.
%
These investigations merely focus on the wind speed and thus do not regard the actual vector, i.e. the veering of
the wind with height is not described and there is little knowledge on the profile of the span-wise velocity component
and the precise shape of the hodograph in the limit of a truly neutral Ekman boundary layer.
%
While this is indeed an academic problem, the Ekman boundary layer often serves as limit for stratified boundary layers
and thus not only needs to be understood but also well-described as reference for higher-order approaches that take into
account possible effects of stratification, roughness or other physical complications of the geophysical system.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Formulation}
%
We consider here incompressible, turbulent Ekman flow, that is, the turbulent flow over a flat rotating plate, as
a physical model for the truly neutral atmospheric boundary layer (ABL).
%
The f-plane approximation is applied such that rotation only acts on horizontal velocity components;
we thus neglect rotational effects on the horizontal components of velocity and dynamical effects
due to latitudinal variation of the rate of rotation.
%
\par
%
The dimensional velocity vector of the numerical simulations is $\underline{U} = (U_1,U_2,U_3) = (U,V,W)$ over the coordinate system
$Oxyz$, where an approximate alignment (plus/minus few degrees centigrade) of the direction $Ox$ with the surface shear stress is achieved.
The coordinate $O_y$ points away from the wall,  and $Oz$ points in the span-wise direction normal to $Oxy$. 
%
For analysis of the results, we use two coordinate systems that are
(i)  exactly aligned with the surface shear stress and labelled by an upper index $\alpha$ as in $\underline{U}^\alpha$ for the velocity vector, and
(ii) the coordinate system aligned with the free-atmosphere geostrophic wind labelled by an upper index $G$ as in $\underline{U}^G$.
%
We denote by $u_\star$ the modulus surface friction velocity and let $Z_\star=1/u_\star$ for brevity.

\section{Towards a Universal Velocity Profile for the Turbulent Ekman Layer}

\end{itemize}
\begin{figure}
  \centerline{\includegraphics[width=0.7\textwidth]{../plot/ustar_alpha.pdf}}
  \caption{Variation of $Z_\star$ and surface veering with Reynolds number according to
    the theory by Spalart et al. (1989) and as estimated from DNS data}
  \label{fig:drag_law}
\end{figure}
% 
\subsection{Drag-Law}
Formulation of a drag-law for Ekman flow consists estimation of two key parameters in the flow as a function of Reynolds number.
%
First, the modulus of the surface friction normalized by the geostrophic wind forcing, $u_\star$ and,
second, the direction of surface shear stress, $\alpha$, also termed wind veering. 
%
A non-zero veering of the wind is a rather special case in comparison with most turbulent flows conisdered in an
engineering context, and it confronts us with a situation where the most appropriate coordinate system for analysis
(namely that aligned with the surface shear stress) is a priori unknown. 
%
We compare our DNS data against a semi-empirical based on integral consideration \cite{Spalart:1989p2436}
and find, as demonstrated in previous work \cite{Ansorge:2014hf} excellent agreement in the range
 $400<\mathbf{Re}<1600$, representing a factor of 16 in variation of viscosity. 
%
\todo{Add LES points}
% 
\par
%
We also find that the solution of the transient equation involved in estimation of $u_\star$ for a given Reynolds number
$Re_D$ is approximated reasonably by the formulation
\begin{equation}
  u_\star= 4\log(\mathbf{Re}_D))-8
\end{equation}
which quantifies the 'weak' dependence of $u_\star$ on the Reynolds number as an approximately logarithmic one.
%

\subsection{Stream-wise Velocity Component} 
For the stream-wise velocity profile (that in non-rotating flows due to the geometry is always aligned with the surface shear stress),
well-established theories exist for various regimes according to their distance from the wall and the relative role of
viscosity, turbulence and interaction with the outer region of the flow with the logarithmic law for the mean velocity as a central anchor point:
\begin{subequations}
  \begin{itemize}
  \item In immediate vicinity to the surface, local turbulent mixing cannot occur for the no-slip/no-penetration boundary condition,
    and the mean velocity is described by a viscous profile of the form
    \begin{equation}
      (U,W)^{\alpha+}  = (z^+,0)  
    \end{equation}
    where the direction of 
    in absence of roughness elements are marginal linear regime / viscous sub-layer \cite{Foken:2002gv,Foken:1978bc}.
    This viscous law has no degree of freedom given the drag-law defining surface friction and veering.
    %
    Theoretical foundation is lacking for the exact shape of the velocity profile in the buffer layer; though crucial for turbulence
    production, it is commonly understood as a transition region between the linear profile at the surface and the logarithmic profile
    aloft.
    A pure blending from the linear velocity profile into the logarithmic one is, however, not reasonable as both the linear and logarithmic
    profile overestimate the velocity in the buffer layer. We therefore introduce a higher-order compensating term into
    the linear profile before blending it into the logarithmic law:
    \begin{equation}
      U_\text{inner}^{\alpha}^+ = \frac{ y^+ + \gamma_4 (y^+)^{4} + \gamma_6 (y^+)^{6} } { 1 + \gamma_6 /u_{ref} (y^+ )^6}, 
    \end{equation} 
    where the fourth- and sixth-order terms in the nominator are designed to fit the profile below $y^+\approx 10$ and
    the sixth-order correction in the denominator limits the growth of $u_\text{inner}$ for large $y^+$. 
  \item In the logarithmic region, we use the profile
    \begin{equation}
      U_\text{log}^{\alpha}^+ = \frac{1}{\kappa} \log y^+ + C 
    \end{equation}
    with $\kappa=0.416$ and $C=5.4605$. %  log-layer (K\'arm\'an, Prandtl, Jimenez, Marusic)
  \item The super-geostrophic maximum of streamwise velocity and a secondary minimum aloft the bulk-turbulent part of
    the boundary layer are well-described by a classic Ekman spiral with adapted boundary conditions and a shift in reference height. 
  \end{itemize}
\end{subequations} 
%
\begin{figure}
  \includegraphics[width=\textwidth]{../plot/u_profile.pdf}
  \caption{Shear-aligned profiles of velocity components $U^{\alpha+}$ in inner (left) and outer (right) units.} 
\end{figure}
